---
title: "hw03-gapminder"
output:
  github_document
---

# Using dplyr, ggplot2 to explore data

## Initialize the data

* Load the gapminder and tidyverse libraries:

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```

* Take a snapshot of the data to *sanity check* that the data and variables appear as we expect:

```{r}
(head(gapminder))
```

*Observations:*
*Using the suppress messages command from cm008 blocks the library loading message from outputting.*

## Task 1

**Task:** Get the max and min GDP per capita for all continents

1. Filter the data by continent
2. Select the GDP per capita
3. Find the minimum and maximum values by: (i) analyzing the range; (ii) querying the min and max directly
4. Present the data in a table
5. Visualize the data in a graph
6. Interpret the data
*Why would we want to look at the max and min GDP per capita for each continent? 
(ii) Imagine we want to look at the distribution of min values and max values respectively to see if they are normal distributions.
(iii) Perhaps we want to see the average min GDP and the average max GDP. Or find the countries with the biggest gap between min and max values, respectively, Or the widest range between min and max values.
*Reflections:*
*Finding the min and max directly is useful if you need to assign the values to individual variables, but using the range() command allows you to see both values faster with a single command. Both methods are used here because their redundancy permits error checking.*
*I used Stack Overflow for some ideas*
*Other resources: https://www.math.ucla.edu/~anderson/rw1001/library/base/html/merge.html; https://www.sixhat.net/how-to-plot-multpile-data-series-with-ggplot.html;
https://stackoverflow.com/questions/23635662/editing-legend-text-labels-in-ggplot*
*Might be more realistic to look only at a particular year, using filter()*
*Make observations about spread, range, thresholds)*
*Used a bar instead of a bar, since this feels like a more natural symbol choice for a upper and lower limit*


```{r}
# Select the necessary data and group to reduce the size; Find the max and min values per continent
min_gdp <- select(gapminder, continent, gdpPercap) %>% 
  group_by(continent) %>% 
    summarize(min_gdp = min(gdpPercap))


max_gdp <- select(gapminder, continent, gdpPercap) %>%
 group_by(continent) %>%
   summarize(max_gdp = max(gdpPercap))

#Merge data into a single table
(min_max_gdp <- merge(min_gdp, max_gdp, by.x = "continent"))

#Sanity check the value
str(min_max_gdp)
(range(gapminder$gdpPercap))

#Visualize the data in a graph
ggplot(min_max_gdp, aes(x = continent, y = value, color = variable)) +
  geom_point(aes(y = min_gdp, col = "min_gdp"), size=20, shape = "-") +
  geom_point(aes(y = max_gdp, col = "max_gdp"), size=20, shape = "-") +
  #Change y-axis to log-scale given the wide range in values
  #scale_y_log10() +
  #Add labels
  labs(title = "Minimum & Maximum GDP per capita, all years",
  x = "Continent", y = "GDP per capita ($)", color = "Legend\n") +
  scale_color_manual(labels = c("Maximum", "Minimum"), values = c("green", "red"))

#Another way to interpret the data
# gapminder %>% 
#   select(year, continent, country, gdpPercap) %>% 
#       group_by(continent) %>% 
#         group_by(year) %>% 
#          summarize(min_gdp = min(gdpPercap))
#ggplot(min_max_gdp, aes(gdpPercap)) +
#  facet_wrap( ~ continent, scale = "free_x") +
#  geom_histogram()

```

## Task 2

**Task:** Look at the spread of GDP per capita within the continents

The initial approach is similar to Task 1:
*1. Filter the data by continent*
*2. Select the GDP per capita*
Now, let's consider the spread of the data, rather than the upper and lower limits:
3. Find the spread of the data
4. Present the data in a table
5. Visualize the data in a graph
6. Interpret the data
(i) Say we want to compare the relative ranges. We could compare just the min and max, but we will get a more complete picture by looking at the distribution.
(ii) Perhaps we want to know which continent has the widest spread? Narrowest spread?
(iii) Referred to 
*Reflections:*
*What different observations would I make based on this data vs. the previous? What are the adv/disadv of this analysis?*
https://www.dummies.com/programming/r/how-to-check-quantiles-in-r/
*the min and max values determined in Task 1 are also shown in the spread, but the spread also gives us more information*
*Could also filter for a specific year or range of years; could also use faceting to display min/max per continent for all years*

```{r}
# Select the necessary data and group to reduce the size; Find the spread of values per continent
#Evaluate the data range using statistical commands
gapminder %>% 
  select(continent, gdpPercap) %>% 
  group_by(continent) %>% 
    summarize(Mean=mean(gdpPercap), SD=sd(gdpPercap), Var=var(gdpPercap))

#Visualize the data in a graph (violin plot)
gapminder %>% 
  select(continent, gdpPercap) %>% 
  group_by(continent) %>% 
    ggplot(aes(continent, gdpPercap)) +
    geom_violin() +
  #Add labels
  labs(title = "Spread of GDP per capita, all years",
  x = "Continent", y = "GDP per capita ($)")

#Visualize the data in a graph (box plot)
gapminder %>% 
  select(continent, gdpPercap) %>% 
  group_by(continent) %>% 
    ggplot(aes(continent, gdpPercap)) +
    geom_boxplot() +
  #Add labels
  labs(title = "Spread of GDP per capita, all years",
  x = "Continent", y = "GDP per capita ($)")
```


## Task 3

**Task:** *Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.*
* I will examine the weighted mean (by population) of life expectancy for different years for Canada

1. Filter the data for Canada
2. Select the year, population and life expectancy
3. x = year, y = life_Exp; weight = pop
4. Present the data in a table
5. Visualize the data in a graph
6. Interpret the data
*Reflections:*

```{r}
gapminder %>% 
  select(year, continent, lifeExp, pop) %>% 
    group_by(year, continent) %>% 
      summarize(mean_lifeExp = mean(lifeExp))
           
```


## Task 4

**Task:** How is life expectancy changing over time on different continents?

1. Filter the data by continent
2. Select the GDP per capita
3. Find the minimum and maximum values by: (i) analyzing the range; (ii) querying the min and max directly
4. Present the data in a table
5. Visualize the data in a graph
6. Interpret the data
*Reflections:*

```{r}

```

